|

### 2021-08-16 dissertation/msc_dissertation.Rmd
adding volcano plot

---
title: |
  | __How analysis strategy affects analysis results__
subtitle: |
  | *Defining the results space in @silberzahn2018 through model specification*
  |
  | Master's thesis - DRAFT
author: "Sebastian Ploner, sploner1@sheffield.ac.uk, University of Sheffield"
linestretch: 1.5
output:
    pdf_document:
        includes:
            in_header: header.tex
        fig_caption: yes
urlcolor: blue
bibliography: references.bib
csl: apa.csl
header-includes:
  - \usepackage[ruled,vlined]{algorithm2e}
  - \usepackage{float}
---

@@

# Introduction
     The Covid-19 pandemic has reaffirmed the need for sound scientific research to inform decision-making on a scientific and societal level [@collins2021]. Rigorous research builds upon a systematic and well-reasoned approach to solving a research problem. Based on available literature, a falsifiable research question is defined and a corresponding hypothesis is developed. An appropriate study is then designed and conducted. To draw inferences from the gathered data, statistical models are developed and the effect of each variable is assessed. Every one of these steps is influenced by the decisions researchers make, which are known as *researcher degrees of freedom* [RDF, @simmons2011; @wicherts2016]. Usually there are many feasible analytical strategies to answering a research question and none of them are inherently right or wrong  [@carp2012]. This often creates uncertainty, for example, about what covariates to include and how to model them, which in turn leads to inconsistent findings [@patel2015; @ioannidis2008]. Recently, efforts have been made to better understand the scope of variation induced by different analytical strategies.

 These efforts include a crowdsourcing approach to data analysis [@silberzahn2018; @botvinik-nezer2020]. Its premise is to have a large number of researchers team up into to smaller, independent groups to investigate the same research question based on the same dataset. For instance, @silberzahn2018 had 29 teams investigate the effects of a football player's skin colour on the odds of being sent off the field. The variation between the analytical strategies was substantial. There were 29 different analyses with 21 different combinations of covariates. Twenty teams found a statistically significant effect. The authors also controlled for researchers' prior believes and experience as well as peer-rated analysis quality, none of which accounted for the variation of results. @botvinik-nezer2020 made similar observations. Crowdsourcing data analysis excels at emphasising the substantial impact of different analytical strategies, but has a significant drawback. It is extremely time and personnel intensive, the two mentioned studies lasted between 2 to 3+ years, and included 61 and 180 analysts, respectively. Not to mention the organisational effort.

 @silberzahn2018 and @botvinik-nezer2020, therefore, suggest using multiverse analysis (also known as specification-curve analysis) as an alternative to crowdsourcing. This approach requires identifying and running all plausible analytical strategies. Plausible strategies are defined as statistically valid, non-redundant, tests appropriate to the research question. Their aggregated results are then used to make inferences about the research question. Despite being statistically more complex and computationally more intense, this approach has the advantages of only needing one, or ideally a couple, researchers. Moreover, a researcher's inherent strategy bias is neutralised and noise is made transparent [@simonsohn2020]. Multiverse approaches are therefore well suited for assessing the scope of variation induced by different analytical strategies. For instance, @patel2015 assessed an effect's robustness by running every combination of covariates and modelling options. Their conclusion was the larger the variation of outcomes, the less robust is the effect, and the less it should therefore to be trusted. The authors termed this measure the *vibration of effect* (VoE).

 Taken together, researchers make many analytical decisions throughout conduction a study which are know as *researcher degrees of freedom*. Crowdsourcing analysis presents a strategy to understand their induced variation, but is impractical due to its substantial time and personnel requirements. Multiverse analysis present a promising, low-personnel, but computationally intense alternative to assessing the scope this variation. Despite observing substantial variation in crowdsourced data analysis projects little is known about the extend to which these approaches cover the full range of possible results. In this project, using principles of multiverse analysis I will define the results space of @silberzahn2018. By doing so I aim at further understanding how a specific analysis strategy affects its result. To my knowldge this is the first project runing not only all *plausible* but all *possible* analytical strategies to a more granular view of the results space.

@@

## statistics are still missing (odds ratio and significance)
id_oddsratio_below_alpha = analysed_specifications %>% select(id, estimate_oddsratio, below_alpha)

@@

# Covariates should be sorted based in their impact.
dat_covariates = dat_covariates[order(dat_covariates$estimate_oddsratio), ]
ordered_covar = dat_covariates$impact_names

# build plot
top = analysed_specifications %>% 
  filter(estimate_oddsratio < 4) %>% 
  ggplot(aes(as.numeric(id), estimate_oddsratio)) +
  geom_point(aes(colour = below_alpha), size = 0.3, alpha = 0.5) +
  geom_ribbon(aes(ymin = ci_lower_oddsratio, ymax = ci_upper_oddsratio), alpha = 0.3) +
  geom_hline(yintercept = 1, color = "black") +
  scale_color_manual(values = c("red", "black")) +
  scale_x_discrete(name = "", expand = c(0.01, 0)) +
  scale_y_continuous(name = "Odds ratio", breaks = c(seq(0.9, 1.7, 0.1)), limits = c(0.89, 1.71), expand = c(0, 0)) +
  labs(title = "Specification curve",
       subtitle = paste0("N = ", nrow(analysed_specifications), ", 95% CI", collapse = "")) +
  theme_classic() +
  theme(panel.grid.major.y = element_line(colour = "grey"),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.line.x = element_blank(),
        plot.margin = unit(c(0, 1, 0, 1), "cm"),
        legend.position = "top",
        legend.title.align = 1,
        legend.text = element_text(size = 11),
        legend.direction = "horizontal",
        legend.title = element_blank(),
        legend.margin = margin(t = -0.5, b = -0.2, unit = "cm"),
        legend.box.margin = margin(t = -0.5, b = -0.2, unit = "cm"),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic", margin = margin(0, 0, 20, 0)),
        text = element_text(size = 10)) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 1), reverse = TRUE))

@@

vibration_of_effect = dat_models_specifications %>%
  filter(estimate_oddsratio < 4) %>% 
  ggplot(aes(x = "", y = estimate_oddsratio)) +
  geom_flat_violin(aes(fill = ""), trim = FALSE, colour = "dark grey", fill = "dark grey") +
  geom_point(aes(x = 0.6, y = estimate_oddsratio, colour = below_alpha),
             position = position_jitter(width = .03, seed = 123), size = 3, shape = 20, alpha = 0.3, stroke = 0) +
  geom_boxplot(aes(x = 0.81, y = estimate_oddsratio), alpha = 0.5, width = 0.1, colour = "black") +
  geom_hline(yintercept = 1.31, linetype = "dashed", color = "black") +
  annotate("text", x = 1.516, y = 1.375, label = "Silberzahn et al. (2018) \nMedian OR", color = "black") +
  scale_color_manual(values = c("red", "black")) +
  scale_y_continuous(name = "Odds ratio", breaks = c(seq(1.0, 1.5, 0.05)), limits = c(0.99, 1.51), expand = c(0, 0)) +
  labs(title = "Results space defined through covariate specification",
       subtitle = paste0("N = ", nrow(dat_models_specifications), ", Odds ratio (OR)", collapse = "")) +
  coord_flip() +
  theme_classic() +
  theme(axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        plot.margin = unit(c(0.5, 1, 0.5, 1), "cm"),
        legend.position = "top",
        legend.justification = "left",
        legend.title.align = 1,
        legend.text = element_text(size = 11),
        legend.direction = "vertical",
        legend.title = element_blank(),
        legend.margin = margin(b = -0.8, unit = "cm"),
        legend.box.margin = margin(b = -0.8, unit = "cm"),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic"),
        text = element_text(size = 10)) +
  guides(colour = guide_legend(override.aes = list(alpha = 1), reverse = TRUE))

@@

## Specified covariate effects
     The final graph shows the specified covariate effects. On the horizontal axis are the covariates, on the vertical axis are the estimates. The covariates are sorted from negative to positive impact. Black indicates a significant effect, red non-significant. The error bars represent the 95% CI. While four covariates are significant, their effects seem relatively weak, ranging from -0.1 to 0.05, only. Interesting is also which covariates are significant. The *team 23* analysis included league country, position, referee and player as covariates.  \newline

@@

covariate_effects = dat_covariates %>%
  ggplot(aes(x = reorder(x = impact_names, X = impact_coef), impact_coef)) +
  geom_point(aes(color = below_alpha)) +
  geom_hline(yintercept = 0) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper, color = below_alpha), width = 0.1) +
  scale_color_manual(values = c("red", "black")) +
  scale_y_continuous(name = "Estimates",
                     labels = scales::comma_format(accuracy = 0.05), breaks = c(seq(-0.15, 0.15, 0.05)), limits = c(-0.15, 0.15)) +
  labs(title = "Specified covariate effects",
       subtitle = paste0("N = ", nrow(analysed_specifications), ", 95% CI", collapse = ""), x = "Covariates") +
  theme_classic() +
  theme(panel.grid.major.y = element_line(colour = "grey"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.margin = unit(c(0.5, 1, 0.5, 1), "cm"),
        legend.position = "top",
        legend.justification = "left",
        legend.title.align = 1,
        legend.text = element_text(size = 11),
        legend.direction = "vertical",
        legend.title = element_blank(),
        legend.margin = margin(l = 0.1, b = -0.729, unit = "cm"),
        legend.box.margin = margin(l = 0.1, b = -0.729, unit = "cm"),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic"),
        text = element_text(size = 10)) +
  guides(colour = guide_legend(reverse = TRUE))

@@

@@

# Discussion

>>>>

---
title: |
  | __How analysis strategy affects analysis results__
subtitle: |
  | *Defining the results space in @silberzahn2018 through model specification*
  |
  | Master's thesis - DRAFT
author: "Sebastian Ploner, sploner1@sheffield.ac.uk, University of Sheffield"
linestretch: 1.5
output:
    pdf_document:
        includes:
            in_header: header.tex
        fig_caption: yes
urlcolor: blue
bibliography: references.bib
csl: apa.csl
header-includes:
  - \usepackage[ruled,vlined]{algorithm2e}
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font={stretch=1}}
---

@@

# Introduction
## Theoretical framework
     The Covid-19 pandemic has reaffirmed the need for sound scientific research to inform decision-making on a scientific and societal level [@collins2021]. Rigorous research builds upon a systematic and well-reasoned approach to solving a research problem. Based on available literature, a falsifiable research question is defined and a corresponding hypothesis is developed. An appropriate study is then designed and conducted. To draw inferences from the gathered data, statistical models are developed and the effect of each variable is assessed. Every one of these steps is influenced by the decisions researchers make, which are known as *researcher degrees of freedom* [RDF, @simmons2011; @wicherts2016]. Usually there are many feasible analytical strategies to answering a research question and none of them are inherently right or wrong  [@carp2012]. This often creates uncertainty, for example, about what covariates to include and how to model them, which in turn leads to inconsistent findings [@patel2015; @ioannidis2008]. Recently, efforts have been made to better understand the scope of variation induced by different analytical strategies.

 These efforts include crowdsourcing approaches to data analysis [@silberzahn2018; @botvinik-nezer2020]. Its premise is to have a large number of researchers team up into to smaller, independent groups to investigate the same research question based on the same dataset. For instance, @silberzahn2018 had 29 teams investigate the effects of a football player's skin colour on the odds of being sent off the field. The variation between the analytical strategies was substantial. There were 29 different analyses with 21 different combinations of covariates. Twenty teams found a statistically significant effect. The authors also controlled for researchers' prior believes and experience as well as peer-rated analysis quality, none of which accounted for the variation of results. @botvinik-nezer2020 made similar observations. Crowdsourcing projects excels at emphasising the substantial impact of different analytical strategies, but they are also extremely time and personnel intensive. The two mentioned studies lasted between 2 to 3+ years, and included 61 and 180 analysts, respectively. Additionally, despite covering a more diverse set of analytical strategies than a single conventional analysis, crowdsourcing is limited by the number of participating teams. It, therefore, still only covers a few selected strategies, not all possible ones.

 @silberzahn2018 and @botvinik-nezer2020, therefore, suggest using multiverse analysis (also known as specification-curve analysis) as an alternative to crowdsourcing. This approach requires identifying and running all plausible analytical strategies. Plausible strategies are defined as statistically valid, non-redundant, tests appropriate to the research question. Their aggregated results are then used to make inferences about the research question. Despite being statistically more complex and computationally more intense, this approach has the advantages of only needing one, or ideally a couple, researchers. Moreover, a researcher's inherent strategy bias is neutralised and noise is made transparent [@simonsohn2020]. Multiverse approaches are therefore well suited for assessing the scope of variation induced by different analytical strategies. However, one aspect has been, to my knowledge, neglected thus far. To properly assess the scope of variation it needs to be contextualised. That is, the space in which it can vary needs to be defined in order to understand to what degree the scope covers the space. The concept of statistical association provides a suitable analogy. Covariation and correlation both indicate the direction of relationship (positive, null, negative), however, only correlation allows for assessing the association's strength. Correlation is the standardised version of covariation, and only knowing the its limits gives the correlation coefficient meaning. The same principle applies to analyses on a broader level.

## Current project
     Taken together, researchers make many analytical decisions throughout conducting a study introducing so-called *researcher degrees of freedom*. Crowdsourcing analysis presents a strategy to understand their induced variation, but is impractical due to its time and personnel requirements, and is still limited by the number of participating teams. Multiverse analysis present a promising, low-personnel, but computationally intense alternative to assessing the scope this variation. Despite observing substantial variation in crowdsourced data analysis projects, little is known about the extend to which these approaches cover the full range of possible results. Determining the full results space and its underlying mechanisms is an important exercise to understand the impact of analytical strategies. To my knowledge no previous study has attempted to define and understand a study's entire results space. In this project, running all possible analytical strategies, I, therefore, aim to define and understand the results space of @silberzahn2018. By doing so, I hope further the understanding of how an analytic strategy affects its result.

@@

## statistics are still missing (odds ratio and significance)
id_oddsratio_below_alpha = analysed_specifications %>% select(id, estimate_oddsratio, below_alpha, p_value)

@@

# Covariates should be sorted based in their impact
dat_covariates = dat_covariates[order(dat_covariates$estimate_oddsratio), ]
ordered_covar = dat_covariates$impact_names

# build plot
top = analysed_specifications %>% 
  filter(estimate_oddsratio < 4) %>% 
  ggplot(aes(as.numeric(id), estimate_oddsratio)) +
  geom_point(aes(colour = below_alpha), size = 0.3, alpha = 0.5) +
  geom_ribbon(aes(ymin = ci_lower_oddsratio, ymax = ci_upper_oddsratio), alpha = 0.3) +
  geom_hline(yintercept = 1, color = "black") +
  scale_color_manual(values = c("red", "black")) +
  scale_x_discrete(name = "", expand = c(0.01, 0)) +
  scale_y_continuous(name = "Odds ratio", breaks = c(seq(0.9, 1.7, 0.1)), limits = c(0.89, 1.71), expand = c(0, 0)) +
  labs(title = "Specification curve",
       subtitle = paste0("N = ", nrow(analysed_specifications), ", 95% CI", collapse = "")) +
  theme_classic() +
  theme(panel.grid.major.y = element_line(colour = "grey"),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.line.x = element_blank(),
        plot.margin = unit(c(0, 1, 0, 1), "cm"),
        legend.position = "top",
        legend.justification = "right",
        legend.title.align = 1,
        legend.text = element_text(size = 10),
        legend.direction = "horizontal",
        legend.title = element_blank(),
        legend.margin = margin(t = -0.61, b = -0.2, unit = "cm"),
        legend.box.margin = margin(t = -0.61, b = -0.2, unit = "cm"),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic", margin = margin(0, 0, 20, 0)),
        text = element_text(size = 10)) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 1), reverse = TRUE))

@@

vibration_of_effect = dat_models_specifications %>%
  filter(estimate_oddsratio < 4) %>% 
  ggplot(aes(x = "", y = estimate_oddsratio)) +
  geom_flat_violin(aes(fill = ""), trim = FALSE, colour = "dark grey", fill = "dark grey") +
  geom_point(aes(x = 0.6, y = estimate_oddsratio, colour = below_alpha),
             position = position_jitter(width = .03, seed = 123), size = 3, shape = 20, alpha = 0.3, stroke = 0) +
  geom_boxplot(aes(x = 0.81, y = estimate_oddsratio), alpha = 0.5, width = 0.1, colour = "black") +
  geom_hline(yintercept = 1.31, linetype = "dashed", color = "black") +
  annotate("text", x = 1.559, y = 1.395, label = "Silberzahn et al. (2018) median OR", color = "black", size = 3.5) +
  scale_color_manual(values = c("red", "black")) +
  scale_y_continuous(name = "Odds ratio", breaks = c(seq(1.0, 1.5, 0.05)), limits = c(0.99, 1.51), expand = c(0, 0)) +
  labs(title = "Results space defined through covariate specification",
       subtitle = paste0("N = ", nrow(dat_models_specifications), ", Odds ratio (OR)", collapse = "")) +
  coord_flip() +
  theme_classic() +
  theme(axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        plot.margin = unit(c(0.5, 1, 0.5, 1), "cm"),
        legend.position = "top",
        legend.justification = "left",
        legend.title.align = 1,
        legend.text = element_text(size = 10),
        legend.direction = "vertical",
        legend.title = element_blank(),
        legend.margin = margin(b = -0.8, unit = "cm"),
        legend.box.margin = margin(b = -0.8, unit = "cm"),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic"),
        text = element_text(size = 10)) +
  guides(colour = guide_legend(override.aes = list(alpha = 1), reverse = TRUE))

@@

## Specified covariate effects
     The final graph shows the specified covariate effects. On the horizontal axis are the covariates, on the vertical axis the estimates. The covariates are sorted from negative to positive impact. Black indicates a significant effect, red non-significant. The error bars represent the 95% CI. Four out of 15 covariates are significant, roughly 27%. Thought the significant covariates' effects seem relatively weak, ranging from -0.1 to 0.05, only. Interesting is also which covariates are significant. The *team 23* analysis included league, position, referee and player as covariates. Of those only player is a significant covariate. Using crude mathematics, one might make the case that the added effect of players is the reason for moving the overall effect from the median, roughly 1.25 to 1.30. Though that's more speculation than causal reasoning.  As mentioned above, the roughly 90% of all outcome were statistically significant, given that only 27% of the covariates have an overall significant effects suggest that the effect is of skin tone itself is fairly robust. \newline

@@

covariate_effects = dat_covariates %>%
  ggplot(aes(x = reorder(x = impact_names, X = impact_coef), impact_coef)) +
  geom_point(aes(color = below_alpha)) +
  geom_hline(yintercept = 0) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper, color = below_alpha), width = 0.1) +
  scale_color_manual(values = c("red", "black")) +
  scale_y_continuous(name = "Estimates",
                     labels = scales::comma_format(accuracy = 0.05), breaks = c(seq(-0.15, 0.15, 0.05)), limits = c(-0.15, 0.15)) +
  labs(title = "Specified covariate effects",
       subtitle = paste0("N = ", nrow(analysed_specifications), ", 95% CI", collapse = ""), x = "Covariates") +
  theme_classic() +
  theme(panel.grid.major.y = element_line(colour = "grey"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.margin = unit(c(0.5, 1, 0.5, 1), "cm"),
        legend.position = "top",
        legend.justification = "right",
        legend.title.align = 1,
        legend.text = element_text(size = 10),
        legend.direction = "horizontal",
        legend.title = element_blank(),
        legend.margin = margin(t = -0.33, b = -0.2, unit = "cm"),
        legend.box.margin = margin(t = -0.33, b = -0.2, unit = "cm"),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic"),
        text = element_text(size = 10)) +
  guides(colour = guide_legend(reverse = TRUE))

@@

```{r volcano_plot, echo = FALSE, fig.width = 8, fig.align = "center"}

volcano_plot = dat_models_specifications %>% 
  filter(estimate_oddsratio < 4) %>% 
  ggplot(aes(estimate_oddsratio, log10(p_value))) +
  geom_point(aes(colour = below_alpha), alpha = 0.3) +
  scale_x_continuous(name = "Odds ratios", breaks = c(seq(1.0, 1.5, 0.05)), limits = c(0.99, 1.51), expand = c(0, 0)) +
  scale_y_continuous(name = "Log10 p-value") +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Vibration of Effect",
       subtitle = paste0("N = ", nrow(analysed_specifications), collapse = "")) +
  theme_classic() +
  theme(panel.grid.major = element_line(colour = "grey"),
        plot.margin = unit(c(0.5, 1, 0.5, 1), "cm"),
        legend.position = "top",
        legend.justification = "right",
        legend.title.align = 1,
        legend.text = element_text(size = 10),
        legend.direction = "horizontal",
        legend.title = element_blank(),
        legend.margin = margin(t = -0.6, b = -0.2, unit = "cm"),
        legend.box.margin = margin(t = -0.6, b = -0.2, unit = "cm"),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic", margin = margin(0, 0, 20, 0)),
        text = element_text(size = 10)) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 1), reverse = TRUE))

volcano_plot

@@

# Discussion
The current project used multiverse analysis to define the results space of the @silberzahn2018. Among others multiverse analysis combines different statistical modelling approaches along with covariate combinations. The goals is assess the robustness of effects. Due to time and resources limitations, I chose to focus on statistical modelling approach, but explore every possible covariate combination. Usually, multiverse analysis would focus on all plausible (or reasonable) analytic strategies. However, this project sought to define the results space of all *possible* analytical outcomes to investigate in which space the outcomes of the *plausible* analytical strategies vary in the first place. Only if the results space is defined, is possible assess its content. The concept of statistical association, provided a neat analogy. There is covariation and correlation, both quantify the variability of two variables. Both indicate the direction of relationship (positive, none, negative). However, to assess the strength of the relationship only correlation is informative. Correlation is the standardised version of covariation, hence, it has a scale to it. And only knowing the limits to space of variation gives the correlation coefficient meaning. The same applies to analyses on a broader level. If researchers want to be able to tell if their analysis reflects a true effect or simply a luck guess, they need to know the entire results space. Only then they can identify the robustness of effects.

 Due to time and resource limitation this project focused on pne

In order to assess the strength of the association covariation is hardly useful as covariate it does not have a standarised scale to it.
